Appendix 3: Estimating species interactions

### 2016-03-14 Harris_2015/manuscript-materials/draft.Rmd
final before resubmission to Ecology?

**Abstract:** Inferring species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that 
outcompetes both of them. Here, I apply models from statistical physics, called 
Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed co-occurrence
rates via maximum likelihood, controlling for indirect effects. Using simulated
landscapes with known interactions, I evaluated Markov networks and six existing
approaches. The Markov networks consistently outperformed other methods, 
correctly isolating direct interactions between species pairs even when indirect
interactions or abiotic factors largely overpowered them. Two computationally 
efficient approximations, which controlled for indirect effects with partial
correlations or generalized linear models, also performed well. Null models 
showed no evidence of being able to control for indirect effects, and reliably 
yielded incorrect inferences when such effects were present.

@@

Below, I demonstrate Markov networks' ability to produce exact predictions
about the direct and indirect consequences of an interaction matrix, and also to
make inferences about the interaction matrix based on co-occurrence rates. Using 
simulated data sets where the "true" interactions are known, I compare this 
approach with several existing methods. Finally, I
discuss opportunities for extending the approach presented here to other
problems in community ecology, e.g. quantifying the overall effect of species
interactions on occurrence rates [@roughgarden_competition_1983] and
disentangling the effects of biotic versus abiotic interactions on species
composition [@pollock_understanding_2014].

@@

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When
Equation 1 can be normalized (Figure 2B), one can find exact maximum likelihood
estimates for $\alpha$ and $\beta$ by numerically optimizing
$p(\vec{y}|\alpha, \beta)$. Fully-observed networks like the ones considered 
here have unimodal likelihood surfaces [@murphy_machine_2012], so these
estimates will represent the global optimum.

@@

I then simulated landscapes with up to 20 interacting species at 25, 200, or
1600 sites using three increasingly complex models (50 replicates for each
combination of size and model). The simplest set of simulated landscapes were
generated with Gibbs samples from Equation 1. For each replicate, I randomly 
drew the “true” $\beta$ coefficient magnitudes from an exponential distribution 
with rate 1 so that most species pairs interacted negligibly but a few 
interactions were strong enough to propagate through the newtwork. I randomly 
assigned 25% of the interactions to be positive; the remainder were negative.

@@

In the final set of landscapes, I simulated each species' abundance (instead of
just presence/absence); furthermore, interactions between species occurred on a 
per-capita basis in these simulations. To prevent runaway mutualisms leading to
infinite abundance, all interaction coefficients were negative in these 
simulations. Good performance on these landscapes would indicate some robustness 
to the mechanistic details of species interactions.

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
models, as described above. For the analyses with 20 species, a 
weakly-informative regularizer (equivalent to a logistic prior with location 0
and scale 2) ensured that the estimates were always finite (Appendix 2).

@@

The third alternative, generalized linear models (GLMs), also provide a
computationally efficient approximation to the Markov network
[@lee_learning_2012].  Following @faisal_inferring_2010,  I fit regularized
logistic regression models [@gelman_weakly_2008] for each species, using the
other species on the landscape as predictors. This produced two interaction
estimates for each species pair; one for the effect of species $i$ on species
$j$ and one for the reverse. These two coefficients are not identifiable from
the data, however [@schmidt_modeling_2012], so I used their average as an
overall measure of the pairwise relationship.

@@

For the null model and the Markov network, the probability of rejecting the
null hypothesis of zero interaction was estimated across a range of "true"
interaction strengths using a generalized additive model with binomial responses
(reject versus not; Appendix 4). The probability of rejection when the "true"
value of $\beta$ was zero was defined as the Type I error rate.

@@

Finally, we can evaluate the models' statistical inferences (focusing on the
first two simulation types, where the interaction coefficients are easiest to
interpret).  The Markov network's Type I error rate depended on the 
simulation type: 0.02 for simulations that matched the model's assumptions,
versus 0.14 for simulations that included environmental heterogeneity (see 
Appendix 4 for confidence interval coverage across a range of $\beta_{ij}$ 
values). The null model's Type I error rates were 0.31 and 0.53 for the
constant and heterogeneous landscapes, respectively---far higher than the 
nominal 0.05 rate. Figure 4C shows, across a range of true interaction 
strengths, the probability that the null model or the Markov network will 
predict the wrong sign of the interaction with 95% confidence.  The null model 
makes such errors more than 5 times as often as the Markov network, even
though it only rejects the null hypothesis twice as often overall (Appendix
4). The Markov network's errors were also more concentrated around 0, and it 
never misclassified strong interactions like the null model did (Figure 4C).

@@

Despite their limitations, Markov networks have enormous potential to improve
ecological understanding. In particular, they make many fewer errors than existing
approaches, and can make precise statements about the conditions where indirect
interactions will overwhelm direct ones. They also provide a simple answer to
the question of how competition should affect a species' overall prevalence,
which has important implications for community-level modeling
[@strong_ecological_1984]. Specifically, Equation 1 can be used to calculate the
expected prevalence of a species in the absence of biotic influences as
$e^{\alpha_i}/(e^{0} + e^{\alpha_i})$. Competition's effect on prevalence can then be
estimated by comparing this value with the observed prevalence (e.g. comparing
Figure 2D with Figure 2C). This novel quantitative result undermines most of
our null models, which unreasonably assume that prevalence would be the exactly
same in the absence of competition as it is in the observed data
[@roughgarden_competition_1983].

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species interactions. Null
and neutral models can be useful for clarifying our thinking
[@harris_occupancy_2011; @xiao_strong_2015], but deviations from a particular
null model must be interpreted with care [@roughgarden_competition_1983]. Even
in small networks with three species, it may simply not be possible to implicate
specific ecological processes like competition by rejecting a general-purpose
null [@gotelli_empirical_2009], especially when the test statistic is
effectively just a correlation coefficient (Figure 4B). When the non-null
backdrop is not controlled for, Type I error rates can skyrocket, the apparent
sign of the interaction can change, and null models can routinely produce
misleading inferences (Figure 1, Figure 4C, @gotelli_empirical_2009).

@@

**Figure 4: A.** The Markov network's estimated interaction
coefficients were generally very similar to the GLM estimates. **B.** The null
model's estimates typically matched the (negative) correlation coefficient,
after controlling for landscape size. **C.** For any given interaction strength,
the null model was much more likely to misclassify its sign with 95% confidence
than the Markov network was.

>>>>

**Abstract:** Inferring species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that 
outcompetes both of them. Here, I apply models from statistical physics, called 
Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed co-occurrence
rates via maximum likelihood, controlling for indirect effects. Using simulated
landscapes with known interactions, I evaluated Markov networks and six existing
approaches. Markov networks consistently outperformed the other methods, 
correctly isolating direct interactions between species pairs even when indirect
interactions or abiotic factors largely overpowered them. Two computationally 
efficient approximations, which controlled for indirect effects with partial
correlations or generalized linear models, also performed well. Null models 
showed no evidence of being able to control for indirect effects, and reliably 
yielded incorrect inferences when such effects were present.

@@

Below, I demonstrate Markov networks' ability to produce exact predictions
about the direct and indirect consequences of an interaction matrix, and also to
make inferences about the interaction matrix based on co-occurrence rates. Using 
simulated data sets where the "true" interactions are known, I compare this 
approach with several existing methods. Finally, I
discuss opportunities for extending the approach presented here to other
problems in community ecology, e.g. quantifying the overall effect of species
interactions on occurrence rates [@roughgarden_competition_1983] and
disentangling the effects of biotic versus abiotic interactions on species
composition [@harris_estimating_2016].

@@

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When
Equation 1 can be normalized (Figure 2B), one can find exact maximum likelihood
estimates for $\alpha$ and $\beta$ by numerically optimizing
$p(\vec{y}|\alpha, \beta)$. Fully-observed networks like the ones considered 
here have unimodal likelihood surfaces [@murphy_machine_2012], so optimizers 
will always find the global optimum.

@@

I then simulated landscapes with up to 20 interacting species at 25, 200, or
1600 sites using three increasingly complex models (50 replicates for each
combination of size and model; Appendix 2). The simplest set of simulated 
landscapes were generated with Gibbs samples from Equation 1. For each 
replicate, I randomly drew the “true” $\beta$ coefficient magnitudes from an 
exponential distribution with rate 1 so that most species pairs interacted 
negligibly but a few interactions were strong enough to propagate through the 
newtwork. I randomly assigned 25% of the interactions to be positive; the 
remainder were negative.

@@

In the final set of landscapes, I simulated each species' abundance (instead of
just presence/absence); furthermore, interactions between species occurred on a 
per-capita basis in these simulations (i.e. each species' effect on the others 
is proportional to its abundance). To prevent runaway mutualisms leading to
infinite abundance, all interaction coefficients were negative in these 
simulations. Good performance on these landscapes would indicate some robustness 
to the mechanistic details of species interactions.

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
models, as described above. For the analyses with 20 species, a 
weakly-informative regularizer (equivalent to a logistic prior with location 0
and scale 2) ensured that the estimates were always finite (Appendix 3).

@@

The third alternative, generalized linear models (GLMs), also provide a
computationally efficient approximation to the Markov network
[@lee_learning_2012].  Following @faisal_inferring_2010, I fit regularized 
logistic regression models [@gelman_weakly_2008] for each 
species, using the other species as predictors. This produced 
two interaction estimates for each species pair (one for the effect of species 
$i$ on species $j$ and one for the reverse). These two estimates were very 
tightly correlated (mean Pearson correlation of 0.95, Appendix 3); their 
arithmetic mean provided a consensus estimate of the overall interaction.

@@

For the null model and the Markov network, the probability of rejecting the
null hypothesis of zero interaction was estimated across a range of "true"
interaction strengths using a kernel smoother (Appendix 4). The probability of 
rejection when the "true" value of $\beta$ was zero was defined as the Type I 
error rate. Because the coefficients' interpretation is different for the 
abundance simulations than for the other two types, its error rates were not
analyzed in this way.

@@

Finally, we can evaluate the models' inferential statistics (focusing on the
first two simulation types, where the interaction coefficients are easiest to
interpret).  The Markov network's Type I error rate was 0.02 for simulations
that matched the model's assumptions, and 0.14 for simulations that included
environmental heterogeneity (see  Appendix 4 for confidence interval coverage
across a range of $\beta_{ij}$ values). In contrast, the null model's Type I 
error rates were 0.30 and 0.51, respectively---far higher than the nominal 0.05 
rate. Figure 4C shows, across a range of true interaction 
strengths, the probability that the null model or the Markov network will 
predict the wrong sign of the interaction with 95% confidence.  The null model 
makes such errors more than 8 times as often as the Markov network, even
though it only rejects the null hypothesis twice as often overall (Appendix
4). The Markov network's errors were also more concentrated around 0, as it 
never misclassified strong interactions like the null model did (Figure 4C).

@@

Markov networks have enormous potential to improve our understanding of species 
interactions. In particular, they make many fewer errors than existing
approaches, and can make precise statements about the conditions where indirect
interactions will overwhelm direct ones. They also provide a simple answer to
the question of how competition should affect a species' overall prevalence,
which has important implications for community-level modeling
[@strong_ecological_1984]. Specifically, Equation 1 can be used to calculate the
expected prevalence of a species in the absence of biotic influences as
$e^{\alpha_i}/(e^{0} + e^{\alpha_i})$. Competition's effect on prevalence can then be
estimated by comparing this value with the observed prevalence (e.g. comparing
Figure 2D with Figure 2C). This novel quantitative result undermines most of
our null models, which unreasonably assume that prevalence would be the exactly
same in the absence of competition as it is in the observed data
[@roughgarden_competition_1983].

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species interactions. 
When the non-null backdrop is not controlled for, Type I error rates can 
skyrocket, the apparent sign of the interaction can change, and null models can 
routinely produce misleading inferences (Figure 1, Figure 4C, 
@gotelli_empirical_2009). Null and neutral models can be useful for clarifying 
our thinking [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a 
given null model must be interpreted with care [@roughgarden_competition_1983]. 
Even in small networks with three species, it may simply not be possible to 
implicate specific ecological processes like competition by rejecting a 
general-purpose null [@gotelli_empirical_2009], especially when the test 
statistic is effectively just a correlation coefficient (Figure 4B).

@@

**Figure 4: A.** The Markov network's estimated interaction
coefficients were generally very similar to the GLM estimates. **B.** The null
model's estimates typically matched the (negative) correlation coefficient,
after controlling for landscape size. **C.** For any given interaction strength,
the null model was much more likely to misclassify its sign with 95% confidence
than the Markov network was. As with the other analyses based on inferential
statistics, this panel only shows data from the first two simulation types.

### 2016-03-14 Harris_2015/manuscript-materials/draft.tex
final before resubmission to Ecology?

\textbf{Abstract:} Inferring species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that outcompetes both of
them. Here, I apply models from statistical physics, called Markov
networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed
co-occurrence rates via maximum likelihood, controlling for indirect
effects. Using simulated landscapes with known interactions, I evaluated
Markov networks and six existing approaches. The Markov networks
consistently outperformed other methods, correctly isolating direct
interactions between species pairs even when indirect interactions or
abiotic factors largely overpowered them. Two computationally efficient
approximations, which controlled for indirect effects with partial
correlations or generalized linear models, also performed well. Null
models showed no evidence of being able to control for indirect effects,
and reliably yielded incorrect inferences when such effects were
present.

@@

Below, I demonstrate Markov networks' ability to produce exact
predictions about the direct and indirect consequences of an interaction
matrix, and also to make inferences about the interaction matrix based
on co-occurrence rates. Using simulated data sets where the ``true''
interactions are known, I compare this approach with several existing
methods. Finally, I discuss opportunities for extending the approach
presented here to other problems in community ecology, e.g.~quantifying
the overall effect of species interactions on occurrence rates
(Roughgarden 1983) and disentangling the effects of biotic versus
abiotic interactions on species composition (Pollock et al. 2014).

@@

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When Equation 1 can be
normalized (Figure 2B), one can find exact maximum likelihood estimates
for \(\alpha\) and \(\beta\) by numerically optimizing
\(p(\vec{y}|\alpha, \beta)\). Fully-observed networks like the ones
considered here have unimodal likelihood surfaces (Murphy 2012), so
these estimates will represent the global optimum.

@@

I then simulated landscapes with up to 20 interacting species at 25,
200, or 1600 sites using three increasingly complex models (50
replicates for each combination of size and model). The simplest set of
simulated landscapes were generated with Gibbs samples from Equation 1.
For each replicate, I randomly drew the ``true'' \(\beta\) coefficient
magnitudes from an exponential distribution with rate 1 so that most
species pairs interacted negligibly but a few interactions were strong
enough to propagate through the newtwork. I randomly assigned 25\% of
the interactions to be positive; the remainder were negative.

@@

In the final set of landscapes, I simulated each species' abundance
(instead of just presence/absence); furthermore, interactions between
species occurred on a per-capita basis in these simulations. To prevent
runaway mutualisms leading to infinite abundance, all interaction
coefficients were negative in these simulations. Good performance on
these landscapes would indicate some robustness to the mechanistic
details of species interactions.

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, a
weakly-informative regularizer (equivalent to a logistic prior with
location 0 and scale 2) ensured that the estimates were always finite
(Appendix 2).

@@

The third alternative, generalized linear models (GLMs), also provide a
computationally efficient approximation to the Markov network (Lee and
Hastie 2012). Following Faisal et al. (2010), I fit regularized logistic
regression models (Gelman et al. 2008) for each species, using the other
species on the landscape as predictors. This produced two interaction
estimates for each species pair; one for the effect of species \(i\) on
species \(j\) and one for the reverse. These two coefficients are not
identifiable from the data, however (Schmidt and Murphy 2012), so I used
their average as an overall measure of the pairwise relationship.

@@

For the null model and the Markov network, the probability of rejecting
the null hypothesis of zero interaction was estimated across a range of
``true'' interaction strengths using a generalized additive model with
binomial responses (reject versus not; Appendix 4). The probability of
rejection when the ``true'' value of \(\beta\) was zero was defined as
the Type I error rate.

@@

Finally, we can evaluate the models' statistical inferences (focusing on
the first two simulation types, where the interaction coefficients are
easiest to interpret). The Markov network's Type I error rate depended
on the simulation type: 0.02 for simulations that matched the model's
assumptions, versus 0.14 for simulations that included environmental
heterogeneity (see Appendix 4 for confidence interval coverage across a
range of \(\beta_{ij}\) values). The null model's Type I error rates
were 0.31 and 0.53 for the constant and heterogeneous landscapes,
respectively---far higher than the nominal 0.05 rate. Figure 4C shows,
across a range of true interaction strengths, the probability that the
null model or the Markov network will predict the wrong sign of the
interaction with 95\% confidence. The null model makes such errors more
than 5 times as often as the Markov network, even though it only rejects
the null hypothesis twice as often overall (Appendix 4). The Markov
network's errors were also more concentrated around 0, and it never
misclassified strong interactions like the null model did (Figure 4C).

@@

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they make many fewer
errors than existing approaches, and can make precise statements about
the conditions where indirect interactions will overwhelm direct ones.
They also provide a simple answer to the question of how competition
should affect a species' overall prevalence, which has important
implications for community-level modeling (Strong et al. 1984).
Specifically, Equation 1 can be used to calculate the expected
prevalence of a species in the absence of biotic influences as
\(e^{\alpha_i}/(e^{0} + e^{\alpha_i})\). Competition's effect on
prevalence can then be estimated by comparing this value with the
observed prevalence (e.g.~comparing Figure 2D with Figure 2C). This
novel quantitative result undermines most of our null models, which
unreasonably assume that prevalence would be the exactly same in the
absence of competition as it is in the observed data (Roughgarden 1983).

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. Null and neutral models can be useful for clarifying our
thinking (Harris et al. 2011, Xiao et al. 2015), but deviations from a
particular null model must be interpreted with care (Roughgarden 1983).
Even in small networks with three species, it may simply not be possible
to implicate specific ecological processes like competition by rejecting
a general-purpose null (Gotelli and Ulrich 2009), especially when the
test statistic is effectively just a correlation coefficient (Figure
4B). When the non-null backdrop is not controlled for, Type I error
rates can skyrocket, the apparent sign of the interaction can change,
and null models can routinely produce misleading inferences (Figure 1,
Figure 4C, Gotelli and Ulrich (2009)).

@@

\textbf{Figure 4: A.} The Markov network's estimated interaction
coefficients were generally very similar to the GLM estimates.
\textbf{B.} The null model's estimates typically matched the (negative)
correlation coefficient, after controlling for landscape size.
\textbf{C.} For any given interaction strength, the null model was much
more likely to misclassify its sign with 95\% confidence than the Markov
network was.

>>>>

\textbf{Abstract:} Inferring species interactions from co-occurrence
data is one of the most controversial tasks in community ecology. One
difficulty is that a single pairwise interaction can ripple through an
ecological network and produce surprising indirect consequences. For
example, the negative correlation between two competing species can be
reversed in the presence of a third species that outcompetes both of
them. Here, I apply models from statistical physics, called Markov
networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observed
co-occurrence rates via maximum likelihood, controlling for indirect
effects. Using simulated landscapes with known interactions, I evaluated
Markov networks and six existing approaches. Markov networks
consistently outperformed the other methods, correctly isolating direct
interactions between species pairs even when indirect interactions or
abiotic factors largely overpowered them. Two computationally efficient
approximations, which controlled for indirect effects with partial
correlations or generalized linear models, also performed well. Null
models showed no evidence of being able to control for indirect effects,
and reliably yielded incorrect inferences when such effects were
present.

@@

Below, I demonstrate Markov networks' ability to produce exact
predictions about the direct and indirect consequences of an interaction
matrix, and also to make inferences about the interaction matrix based
on co-occurrence rates. Using simulated data sets where the ``true''
interactions are known, I compare this approach with several existing
methods. Finally, I discuss opportunities for extending the approach
presented here to other problems in community ecology, e.g.~quantifying
the overall effect of species interactions on occurrence rates
(Roughgarden 1983) and disentangling the effects of biotic versus
abiotic interactions on species composition (Harris 2016).

@@

\noindent \textbf{Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.}
In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a set of independent \(\vec{y}\) vectors indicating which species are
present at each site on the landscape). When Equation 1 can be
normalized (Figure 2B), one can find exact maximum likelihood estimates
for \(\alpha\) and \(\beta\) by numerically optimizing
\(p(\vec{y}|\alpha, \beta)\). Fully-observed networks like the ones
considered here have unimodal likelihood surfaces (Murphy 2012), so
optimizers will always find the global optimum.

@@

I then simulated landscapes with up to 20 interacting species at 25,
200, or 1600 sites using three increasingly complex models (50
replicates for each combination of size and model; Appendix 2). The
simplest set of simulated landscapes were generated with Gibbs samples
from Equation 1. For each replicate, I randomly drew the ``true''
\(\beta\) coefficient magnitudes from an exponential distribution with
rate 1 so that most species pairs interacted negligibly but a few
interactions were strong enough to propagate through the newtwork. I
randomly assigned 25\% of the interactions to be positive; the remainder
were negative.

@@

In the final set of landscapes, I simulated each species' abundance
(instead of just presence/absence); furthermore, interactions between
species occurred on a per-capita basis in these simulations (i.e.~each
species' effect on the others is proportional to its abundance). To
prevent runaway mutualisms leading to infinite abundance, all
interaction coefficients were negative in these simulations. Good
performance on these landscapes would indicate some robustness to the
mechanistic details of species interactions.

\noindent \textbf{Recovering species interactions from simulated data.}
I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov network
models, as described above. For the analyses with 20 species, a
weakly-informative regularizer (equivalent to a logistic prior with
location 0 and scale 2) ensured that the estimates were always finite
(Appendix 3).

@@

The third alternative, generalized linear models (GLMs), also provide a
computationally efficient approximation to the Markov network (Lee and
Hastie 2012). Following Faisal et al. (2010), I fit regularized logistic
regression models (Gelman et al. 2008) for each species, using the other
species as predictors. This produced two interaction estimates for each
species pair (one for the effect of species \(i\) on species \(j\) and
one for the reverse). These two estimates were very tightly correlated
(mean Pearson correlation of 0.95, Appendix 3); their arithmetic mean
provided a consensus estimate of the overall interaction.

@@

For the null model and the Markov network, the probability of rejecting
the null hypothesis of zero interaction was estimated across a range of
``true'' interaction strengths using a kernel smoother (Appendix 4). The
probability of rejection when the ``true'' value of \(\beta\) was zero
was defined as the Type I error rate. Because the coefficients'
interpretation is different for the abundance simulations than for the
other two types, its error rates were not analyzed in this way.

@@

Finally, we can evaluate the models' inferential statistics (focusing on
the first two simulation types, where the interaction coefficients are
easiest to interpret). The Markov network's Type I error rate was 0.02
for simulations that matched the model's assumptions, and 0.14 for
simulations that included environmental heterogeneity (see Appendix 4
for confidence interval coverage across a range of \(\beta_{ij}\)
values). In contrast, the null model's Type I error rates were 0.30 and
0.51, respectively---far higher than the nominal 0.05 rate. Figure 4C
shows, across a range of true interaction strengths, the probability
that the null model or the Markov network will predict the wrong sign of
the interaction with 95\% confidence. The null model makes such errors
more than 8 times as often as the Markov network, even though it only
rejects the null hypothesis twice as often overall (Appendix 4). The
Markov network's errors were also more concentrated around 0, as it
never misclassified strong interactions like the null model did (Figure
4C).

@@

Markov networks have enormous potential to improve our understanding of
species interactions. In particular, they make many fewer errors than
existing approaches, and can make precise statements about the
conditions where indirect interactions will overwhelm direct ones. They
also provide a simple answer to the question of how competition should
affect a species' overall prevalence, which has important implications
for community-level modeling (Strong et al. 1984). Specifically,
Equation 1 can be used to calculate the expected prevalence of a species
in the absence of biotic influences as
\(e^{\alpha_i}/(e^{0} + e^{\alpha_i})\). Competition's effect on
prevalence can then be estimated by comparing this value with the
observed prevalence (e.g.~comparing Figure 2D with Figure 2C). This
novel quantitative result undermines most of our null models, which
unreasonably assume that prevalence would be the exactly same in the
absence of competition as it is in the observed data (Roughgarden 1983).

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. When the non-null backdrop is not controlled for, Type I
error rates can skyrocket, the apparent sign of the interaction can
change, and null models can routinely produce misleading inferences
(Figure 1, Figure 4C, Gotelli and Ulrich (2009)). Null and neutral
models can be useful for clarifying our thinking (Harris et al. 2011,
Xiao et al. 2015), but deviations from a given null model must be
interpreted with care (Roughgarden 1983). Even in small networks with
three species, it may simply not be possible to implicate specific
ecological processes like competition by rejecting a general-purpose
null (Gotelli and Ulrich 2009), especially when the test statistic is
effectively just a correlation coefficient (Figure 4B).

@@

\textbf{Figure 4: A.} The Markov network's estimated interaction
coefficients were generally very similar to the GLM estimates.
\textbf{B.} The null model's estimates typically matched the (negative)
correlation coefficient, after controlling for landscape size.
\textbf{C.} For any given interaction strength, the null model was much
more likely to misclassify its sign with 95\% confidence than the Markov
network was. As with the other analyses based on inferential statistics,
this panel only shows data from the first two simulation types.

### 2016-03-14 Harris_2015/recovery.Rmd
final before resubmission to Ecology?

This document describes how the different models were fit to the simulated data 
from Appendix 2 and how each model's performance was evaluated.[^1]

@@

```{r, cache = FALSE, eval = FALSE}

# Find all the csv files in the fakedata/matrices folder,
# then drop .csv
identifiers = dir("fakedata/matrices", pattern = "\\.csv$") %>%
  gsub("\\.csv$", "", .)

mclapply(identifiers, fit_all, mc.cores = mc.cores, 
         mc.preschedule = FALSE, mc.silent = TRUE) %>% 
  bind_rows() %>% 
  write.csv(file = "estimates.csv")
```

### Logistic prior used for the Markov network

@@

>>>>

This document describes how the different models were fit to the simulated data 
from Appendix 2 and how each model's performance was evaluated.

@@

```{r, cache = FALSE, eval = TRUE}

# Find all the csv files in the fakedata/matrices folder,
# then drop .csv
identifiers = dir("fakedata/matrices", pattern = "\\.csv$") %>%
  gsub("\\.csv$", "", .)
```

```{r, eval = FALSE}
mclapply(identifiers, fit_all, mc.cores = mc.cores, 
         mc.preschedule = FALSE, mc.silent = TRUE) %>% 
  bind_rows() %>% 
  write.csv(file = "estimates.csv")
```

## Logistic prior used for the Markov network

@@

## GLM coefficient correlations

The GLM method produces two estimates for each species pair.  These estimates
tend to be very similar, so most reasonable methods for generating a consensus
estimate should produce very similar results.

```{r, cache=TRUE}
get_glm_cor = function(identifier){
  data_filename = paste0("fakedata/matrices/", identifier, ".csv")

  x = raw_obs[ , species_is_variable]

  coef_matrix = matrix(0, ncol(x), ncol(x))

  # Fill in the coefficient matrix, column-by-column
  for(i in 1:ncol(x)){
    if(var(x[,i]) > 0){
      coefs = coef(bayesglm(x[,i] ~ x[ , -i], family = binomial))[-1]
      coef_matrix[i, -i] = coefs
    }
  }

  # Compare the estimates from the upper triangle with the estimates from 
  # the estimates in the lower triangle (which becomes the upper triangle after 
  # transposing)
  cor(coef_matrix[upper.tri(coef_matrix)], t(coef_matrix)[upper.tri(coef_matrix)])
}

glm_correlations = unlist(mclapply(identifiers, get_glm_cor))

mean(glm_correlations)

hist(glm_correlations)
```
